---
title: "eStar package"
output: 
  rmarkdown::html_vignette:
    fig_width: 5
    fig_height: 3
vignette: >
  %\VignetteIndexEntry{eStar_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(eStar)
library(dplyr) ## TODO: shouldn't it be imported by eStar
library(ggplot2)
library(cowplot)
```

# Toy data set
```{r toy-data, fig.height = 6, fig.width = 9, fig.cap = "a) Baseline and response time-series of the state variable created for the toy data set and b) long response ratio of response/baseline values.."}
load("../data/toy_blts.rda")
load("../data/toy_svts.rda")
sv_df <- dplyr::left_join(dplyr::rename(toy_blts, "bl" = stat_var),
          dplyr::rename(toy_svts, "resp" = stat_var),
          by = "time") 

plot_grid(
    sv_df %>%
    tidyr::pivot_longer(cols = c("bl", "resp"), names_to = "data", values_to = "value")%>%
    dplyr::mutate(data = dplyr::case_when(
                                    data == "bl" ~ "Baseline",
                                    data == "resp" ~ "Response")) %>%
    ggplot(aes(x = time, y = value, colour = data, group = data)) +
    geom_line()+
    geom_point()+
    scale_color_viridis_d(option = "cividis")+
    labs(x = "Time", y = "State variable", color = "Data")+
    theme_minimal() +
    theme(legend.position = c(0.80, 0.20),
          legend.background = element_rect(fill="white")),
    sv_df %>%
    dplyr::mutate(lrr = log(resp / bl)) %>%
    ggplot(aes(x = time, y = lrr), colour = "green") +
    geom_line()+
    geom_point()+
    labs(x = "Time", y = "Log response ratio")+
    theme_minimal(),
    nrow = 1,
    labels = c("a)", "b)"))
```

# Main functions
Examples of each function appled to the toy dataset.

## Invariability

Invariability calculated as the inverse of the coefficient of variation of a state 
variable over a defined time frame (`tf_resp`).
```{r}
invariability(
   sv_resp = "stat_var", t_resp = "time", mode = "cv",
   tf_resp = c(11, 50), data_resp = toy_svts  ## V: one point to think about is a suggestion to the user whether the first time step should be AFTER the disturbance (as you selected here, 11. Or the step OF the disturbance. ATM I do not have clear point of view on that)
 )

## quick check
st <- toy_svts$stat_var[11:50]
mean(st) / sd(st) 
## diffeers from the one obtained with the function, identifies that the subsettign of the time serties is not done correctly

invariability(
   sv_resp = toy_svts$stat_var, t_resp = toy_svts$time, mode = "cv",
   tf_resp = c(11, 50)
 )
```
Invariability calculated as the inverse of the residuals of the linear model
fiited to the log response ratio (response/baseline) over time.
```{r}
invariability(
    sv_resp = "stat_var", t_resp = "time", mode = "lm_res",
    tf_resp = c(11, 50), data_resp = toy_svts, sv_bl = "stat_var",
    t_bl = "time", tf_bl = c(11, 50), data_bl = toy_blts
)
invariability(
   sv_resp = toy_svts$stat_var, t_resp = toy_svts$time,
   tf_resp = c(11, 50), mode = "lm_res", sv_bl = toy_blts$stat_var,
   t_bl = toy_blts$time, tf_bl = c(11, 50)
)
```

## Resistance

Resistance calculated in relation to a baseline time series (`bl_mode = "ts"`), 
as the log response ratio at a given time (`res_time = "single"`, `t_res = 11`).
```{r}
resistance(
  sv_resp = "stat_var", t_resp = "time", data_resp = toy_svts, bl_mode = "ts",
  sv_bl = "stat_var", t_bl = "time", data_bl = toy_blts,
  res_time = "single", t_res = 11
)
```

Resistance calculated in relation to a previous valeus of the state variable
(`bl_mode = "point"`, `tresp_bl = 9`), as the log response ratio at a given 
time (`res_time = "single"`, `t_res = 11`).
```{r}
resistance(
  sv_resp = "stat_var", t_resp = "time", data_resp = toy_svts, bl_mode = "point",
  tresp_bl = 9, res_time = "single", t_res = 11
)
```

Resistance calculated in relation to a baseline time series (`bl_mode = "ts"`), 
as the highest log response ratio during a given time frame (`res_time = "time_frame", 
tf_res = c(11, 50)`).
```{r}
resistance(
  sv_resp = "stat_var", t_resp = "time", data_resp = toy_svts, bl_mode = "ts",
  sv_bl = "stat_var", t_bl = "time", data_bl = toy_blts,
  res_time = "time_frame", tf_res = c(11, 50)
)
```

Resistance calculated in relation to a previous value (`bl_mode = "point"`), 
as the highest log response ratio during a given time frame (`res_time = "time_frame", 
tf_res = c(11, 50)`).
```{r}
resistance(
  sv_resp = "stat_var", t_resp = "time", data_resp = toy_svts, bl_mode = "point",
  tresp_bl = 9, res_time = "time_frame", tf_res = c(11, 50)
)
```
## Rate of recovery

Rate of recovery calculated in relation to a baseline (`bl_mode = "ts"`), 
as the slope of a linear model (`slope_mode = "lm"`) of the log response ratio
over a time frame (`tf_slope = c(12, 50)`).
```{r}
recovery_rate(
   sv_resp = "stat_var", t_resp = "time", data_resp = toy_svts, bl_mode = "ts",
   sv_bl = "stat_var", t_bl = "time", data_bl = toy_blts,
   slope_mode = "lm", tf_slope = c(12, 50)
)
```
Rate of recovery calculated in relation to a previous point (`bl_mode = "point"`), 
as the slope of a linear model (`slope_mode = "lm"`) of the log response ratio
over a time frame (`tf_slope = c(12, 50)`).
```{r}
 recovery_rate(
   sv_resp = "stat_var", t_resp = "time", data_resp = toy_svts, bl_mode = "point",
   t_rec = 9, sv_bl = "stat_var", t_bl = "time", data_bl = toy_blts,
   slope_mode = "lm", tf_slope = c(12, 50)
 )
```

Rate of recovery calculated in relation to a baseline (`bl_mode = "ts"`), 
as the slope of the log response ratio between two points (`slope_mode = "points"`)
over a time frame (`tf_slope = c(12, 50)`).
```{r}
 recovery_rate(
   sv_resp = "stat_var", t_resp = "time", data_resp = toy_svts, bl_mode = "ts",
   t_rec = 50, sv_bl = "stat_var", t_bl = "time", data_bl = toy_blts,
   slope_mode = "points", tf_slope = c(12, 50)
 )
```

Rate of recovery calculated in relation to a previous point (`bl_mode = "point"`), 
as the slope of the log response ratio between two points (`slope_mode = "points"`)
over a time frame (`tf_slope = c(12, 50)`).
```{r}
 recovery_rate(
   sv_resp = "stat_var", t_resp = "time", data_resp = toy_svts, bl_mode = "point",
   t_rec = 9, sv_bl = "stat_var", t_bl = "time", data_bl = toy_blts,
   slope_mode = "points", tf_slope = c(12, 50)
 )
```

## Extent of recovery
Extent of recovery as the log response ratio in relation to a baseline time-series
(`bl_mode = "ts"`), a certain time (`t_rec = 50`).
```{r}
recovery_extent(
 sv_resp = "stat_var", t_resp = "time", data_resp = toy_svts, bl_mode = "ts",
 t_rec = 50, sv_bl = "stat_var", t_bl = "time", data_bl = toy_blts
)
```
Extent of recovery as the log response ratio in relation to a previous point in time
(`bl_mode = "point"`), a certain time (`t_rec = 50`).
```{r}
recovery_extent(
  sv_resp = "stat_var", t_resp = "time", data_resp = toy_svts, bl_mode = "point",
  t_rec = 9, sv_bl = "stat_var", t_bl = "time", data_bl = toy_blts
)
```

<details><summary>Session Info</summary>

```{r sessionInfo}
## DO NOT REMOVE!
## We store the settings of your computer and the current versions of the
## packages used to allow for reproducibility
Sys.time()
#git2r::repository() ## uncomment if you are using GitHub
sessionInfo()
```

</details>
