---
title: "MARSS in estar and its multivariate metrics"
output: 
  rmarkdown::html_vignette:
    toc: TRUE
    toc_depth: 2 
    fig_caption: TRUE
    code_folding: hide
vignette: >
  %\VignetteIndexEntry{marss_estar}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
bibliography: references.bib
editor_options: 
  markdown: 
    wrap: sentence
---

```{r knit-setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(
  comment = "#>"
  )
options(rmarkdown.html_vignette.check_title = FALSE) ## title and VignetteIndexEntry are slightly different and this avoids a warning message every time it's rendered
```

```{r message=FALSE}
library(estar)
library(tidyverse)
library(viridis)
library(MARSS)
library(hesim)
source("custom_aesthetics.R")

label_intensity <- function(intensity, mu){
  paste0("Conc = ", intensity, " micro g/L")
}
```

Calculation of the so-called "multivariate stability metrics" (in the context of the `estar` package) relies on the interaction strengths between species composing a community.
This information is classically summarized in the *community matrix* [sensu @novak_characterizing_2016, detailed in the Glossary in the main text].
In community models, this is equivalent to a time-discrete version [@downing_temporal_2020] of the Jacobian matrix composed of the partial derivatives of the dynamical equations describing each species population growth [@dominguez-garcia_unveiling_2019].
In real-world communities, interaction strengths can be estimated with multivariate autoregressive state-space (MARSS) models.

In this vignette, we focus on explaining [how a MARSS model works](#intro-marss-model), how these models can be used with the functions provided by the [`estar` package](#marss-in-estar), and the issues that may arise when empirical data is [limited](#limited-data).
To facilitate understanding, we illustrate the definition of a MARSS model in the context of this application to an [aquatic community](#aquatic-data).
For a more general and formal definition, refer to the [`MARSS` package documentation](https://nwfsc-timeseries.github.io/MARSS/).

# Overview of demonstration data

Similar to what we did for the univariate stability metrics, we illustrate the use of multivariate stability metrics on a dataset compiled from an ecotoxicological study about the effects of insecticide on a community of aquatic macroinvertebrates (references anonymized for review).
The insecticide chlorpyrifos negatively affects freshwater macroinvertebrates, and, to a lesser degree, zooplankton species (references anonymized for review).
This insecticide was applied at four different concentrations (0.1, 0.9, 6, and 44 $\mu$g/ L), with two replicates per concentration level.
Additionally, four replicates were used as control, i.e. an undisturbed systems ('baseline' in `estar` terminology).
The community is composed of 128 species, classified into 5 functional groups: herbivores, detritiherbivores, carnivores, omnivores, and detritivores.

```{r echo = FALSE, message = FALSE,  fig.height = 6, fig.width = 8, fig.cap = "Organism mean count (log axis) of the five functional groups over the course of ecotoxicological experiment. The vertical dashed line identifies the time when chloryfiros insecticide was applied (at week 0), and facets represent the different concentrations (micro g/L)."}
(aquacomm_fgps.logplot <- aquacomm_fgps |>
  tidyr::pivot_longer(cols = c(herb, detr_herb, carn, omni, detr),
                      names_to = "gp", values_to = "abund") |>
  ## summarize to plot
  dplyr::group_by(time, treat, gp) |>
  dplyr::summarize(abund_logmean = log(mean(abund))) |>
  dplyr::ungroup() |>
  dplyr::mutate(gp = forcats::fct_recode(factor(gp), 
                      Herbivore = "herb",
                      Detr_Herbivore = "detr_herb",
                      Carnivore = "carn",
                      Omnivore = "omni",
                      Detritivore = "detr")) |>
  ggplot(aes(x = time, y = abund_logmean, group = gp, colour = gp)) +
  geom_line() +
  geom_point(size = 1) +
  geom_vline(aes(xintercept = 0), linetype = 2, colour = "black") +
  scale_colour_manual(values = intensity_colors[seq(from = 1, by = 2, 
                                                    length.out = 5)]) +
  facet_wrap(~treat, nrow = 5, labeller = labeller(treat = label_intensity)) +
  labs(x = "Time (week)", y = "Mean abundance (log)", 
       colour = "Functional\ngroup") +
  theme_estar())
```

# MARSS models

## Introduction to MARSS models {#intro-marss-model}

A complete MARSS model that describes a state process ($\mathbf{X}$) assessed via an observation process ($\mathbf{Y}$) is defined as: \begin{equation}
\begin{gathered}
\mathbf{x}_t = \mathbf{B}\mathbf{x}_{t-1}+\mathbf{u}+\mathbf{C}_{t}\mathbf{c}_{t}+\mathbf{w}_t \text{ where } \mathbf{w}_t \sim \ \text{MVN}(0,\mathbf{Q}) \\
\mathbf{y}_t = \mathbf{Z}\mathbf{x}_t+\mathbf{a}+\mathbf{D}_{t}\mathbf{d}_{t}+\mathbf{v}_t \text{ where } \mathbf{v}_t \sim \ \text{MVN}(0,\mathbf{R}) \\
\mathbf{x}_1 \sim\ \text{MVN}(\pi,\Lambda) \text{ or } \mathbf{x}_0 \sim\ \text{MVN}(\pi,\Lambda)
\end{gathered}
\end{equation}

where

-   $\mathbf{X}$ is a $m \times T$ matrix, composed of the unknown abundance values of $m$ species or functional groups ($fg$, rows), over time $T$ ($t$, columns)

```{=tex}
\begin{equation}
X=
\begin{bmatrix}
    x_{fg1, t1} & x_{fg1, t2} & ... & x_{fg1, tT}\\
    x_{fg2, t1} & x_{fg2, t2} & ... & x_{fg2, tT}\\
      ... & ... & ... & ...\\
      x_{fgm, t1} & x_{fgm, t2} & ... & x_{fgm, tT}
\end{bmatrix}
\end{equation}
```
-   $\mathbf{Y}$ is a $n \times T$ matrix, composed of $n$ observations (rows) over time $T$ ($t$, columns)

```{=tex}
\begin{equation}
Y=
\begin{bmatrix}
    y_{1, t1} & y_{1, t2} & ... & y_{1, tT}\\
    y_{2, t1} & y_{2, t2} & ... & y_{2, tT}\\
    ... & ... & ... & ...\\
    y_{n, t1} & y_{n, t2} & ... & y_{n, tT}
\end{bmatrix}
\end{equation}
```
Note that:

-   considering that the observations in $\mathbf{Y}$ refer to processes $\mathbf{X}$, $n \geq m$. Missing values can be passed as `NA`.
-   from $\mathbf{X}$ and $\mathbf{Y}$ alone, it is impossible to trace how observations refer to processes. We will address this in the description of $\mathbf{Y}$.

The state process ($\mathbf{X}$) is described by:

-   $\mathbf{x}_t$ is a $m \times 1$ vector of unknown abundances of the $m$ species at time $t$.
-   $\mathbf{B}$ is a $m \times m$ matrix of states covariances. This is the matrix of **interactions strengths** ($\beta$) we want to estimate to calculate community-based stability metrics.
-   $\mathbf{u}$ is a $m \times 1$ vector of states fixed values. For example, expected abundance values for each functional group.
-   $\mathbf{C}_t$ is a $m \times p$ matrix of coefficients ($\gamma$) of each of the $p$ covariates (enviromental conditions) effects on each of the $m$ states (abundances).
-   $\mathbf{c}_t$ is a $p \times 1$ vector of covariate values. For example, environmental conditions which are expected to affect abundances.
-   $\mathbf{w}_t$ is a $m \times 1$ vector of the state ($X$) process errors. These account for the stochasticity of the system.

Thus, in matrix formulation:

```{=tex}
\begin{equation}
\begin{bmatrix}
    x_{1}\\
    ... \\
    x_{m}
    \end{bmatrix}_t = 
\begin{bmatrix}
    \beta_{\mathrm{1,1}} & ... & \beta_{\mathrm{1,m}} \\
    ... & ... & ... \\
    \beta_{\mathrm{m,1}} & ... & \beta_{\mathrm{m,m}}
\end{bmatrix}
\begin{bmatrix}
    x_{1}\\
    ... \\
    x_{m}
\end{bmatrix}_{t-1} 
+
\begin{bmatrix}
    u_{1}\\
    ...\\
    u_{m}
\end{bmatrix} 
+ 
\begin{bmatrix}
    \gamma_{\mathrm{1,1}} & ... & \gamma_{\mathrm{1,p}} \\
    ... & ... & ... \\
    \gamma_{\mathrm{m,1}} & ... & \gamma_{\mathrm{m,p}}
\end{bmatrix}
\begin{bmatrix}
    c_{\mathrm{1}} \\
    ... \\
    c_{\mathrm{m}}
\end{bmatrix}_{t}
+
\begin{bmatrix}
w_{1}\\
...\\
w_{m}
\end{bmatrix}_t \\
\end{equation}
```
The $w_{t}$ errors follow a multivariate normal distribution of mean 0, and covariance $\mathbf{Q}_t$ (a $m \times m$ matrix).
\begin{equation}
\mathbf{w}_t \sim \,\text{MVN}
\begin{pmatrix}
0,
    \begin{bmatrix}
        \sigma_{\mathrm{1}} & ... & \sigma_{\mathrm{1,m}} \\
        ... & ... & ... \\
        \sigma_{\mathrm{m,1}} & ... & \sigma_{\mathrm{m}} \\ 
        \end{bmatrix}
\end{pmatrix}\\
\end{equation}

where ($\sigma$) refer to variances of abundances (diagonal) and covariances among abundances of different species (off-diagonal).

The model initial conditions can be specified at $t=0$ ($x_0$) or $t=1$ ($x_1$).

```{=tex}
\begin{equation}
\begin{bmatrix}
    x_{fg1}\\
    ...\\
    x_{fgm}
\end{bmatrix}_0 
= 
\begin{bmatrix}
    \mu_1\\
    ...\\
    \mu_m
\end{bmatrix}_t
\end{equation}
```
The observation process ($\mathbf{Y}$) is described by:

-   $\mathbf{y}_t$ is a $m \times 1$ vector with the number of $n$ observations taken at time $t$.
-   $\mathbf{Z}_t$ is a $n \times m$ matrix which describes the relationships ($\zeta$) between hidden states $\mathbf{x}$ (real abundances of functional groups) and the observations $\mathbf{y}$ (samples).
-   $\mathbf{a}$ is a $n \times 1$ vector of fixed values.
-   $\mathbf{D}_t$ is a $n \times q$ matrix of $\delta$ coefficients of each of the $q$ covariates on each of the $n$ observations.
-   $\mathbf{d}_t$ s a $q \times 1$ vector of covariate values affecting the observation process.
-   $\mathbf{v}_t$ is a $n \times 1$ vector of observation errors. They follow a multivariate normal distribution with mean 0, and covariance $\mathbf{R}_t$ (a $n \times n$ matrix) .

In matrix formulation: 

```{=tex}
\begin{equation}
\begin{bmatrix}
    y_{1}\\
    ... \\
    y_{n}
    \end{bmatrix}_t = 
\begin{bmatrix}
    \zeta_{\mathrm{1,1}} & ... & \zeta_{\mathrm{1,m}} \\
    ... & ... & ... \\
    \zeta_{\mathrm{m,1}} & ... & \zeta_{\mathrm{m,m}}
\end{bmatrix}
\begin{bmatrix}
    x_{1}\\
    ... \\
    x_{m}
\end{bmatrix}_{t-1} 
+
\begin{bmatrix}
    a_{obs1}\\
    ...\\
    a_{obsn}
\end{bmatrix} 
+
\begin{bmatrix}
    \delta_{\mathrm{1,1}} & ... & \delta_{\mathrm{1,q}} \\
    ... & ... & ... \\
    \delta_{\mathrm{n,1}} & ... & \delta_{\mathrm{q,q}}
\end{bmatrix}
\begin{bmatrix}
    d_{\mathrm{1}} \\
    ... \\
    d_{\mathrm{q}}
\end{bmatrix}_{t}
+
\begin{bmatrix}
v_{obs1}\\
...\\
v_{obsn}
\end{bmatrix}_t \\
\end{equation}
```

The $v_{t}$ errors follow a multivariate normal distribution of mean 0, and covariance $\mathbf{R}_t$ (a $n \times n$ matrix).
\begin{equation}
\mathbf{v}_t \sim \,\text{MVN}
\begin{pmatrix}
0,
    \begin{bmatrix}
        \sigma_{\mathrm{1}} & ... & \sigma_{\mathrm{1,n}} \\
        ... & ... & ... \\
        \sigma_{\mathrm{n,1}} & ... & \sigma_{\mathrm{n}} \\ 
        \end{bmatrix}
\end{pmatrix}\\
\end{equation}

The terms $\mathbf{y}_t$, $\mathbf{c}_t$ and $\mathbf{d}_t$ are empirical data that are provided by the user.
The remaining terms $\mathbf{B}$, $\mathbf{u}$, $\mathbf{C}$, $\mathbf{Q}$, $\mathbf{Z}$, $\mathbf{a}$, $\mathbf{D}$, and $\mathbf{R}$ can be estimated or some of them (e.g. $\mathbf{u}$, $\mathbf{Q}$) can be fixed to certain values based on the assumptions about the system.
The terms $\mathbf{c}_t$ and $\mathbf{d}_t$ must be given as an input, without any missing values.

## MARSS in estar {#marss-in-estar}

When calculating the multivariate metrics the user can either provide the required community matrix ($\mathbf{B}$) or have it estimated from the empirical data using a set of predefined assumptions about the modeled community.
Empirical data in this case constitutes time series of population abundances (or biomass) of all the species (or functional groups) in the community.
In this section we detail what are these assumptions and show how they might affect the estimation of ($\mathbf{B}$) by applying them to the same study case.
None of the parameters detailed here have default values in functions that calculate the community-based stability metrics.
The motivation for it is to avoid the indiscriminate use, and thus force the user to take into consideration the implications of the chosen assumptions.

We start from the minimal case of one single observation per state process and other simplifying assumptions, and build its complexity up it step-wise.

## Minimal case

In our minimal case, we consider that

-   Each observation corresponds to one state process ($y_i = x_i$), thus $\mathbf{Z}$ = "identity". In our case we have 5 functional groups, therefore $\mathbf{Z}$ corresponds to a $5 \times 5$ identity matrix, since $n = m$:

```{r}
Z_I5 <- matrix(list(0), 5, 5)
diag(Z_I5) <- 1
```

-   All functional groups have the same state processes variance\*: $w_t = \sigma$, thus `Q = "diagonal and equal"` (in MARSS terminology), which equates to \begin{equation}
    \begin{bmatrix}
          \sigma & 0 & 0 & 0 & 0 \\
          0 & \sigma & 0 & 0 & 0 \\
          0 & 0 & \sigma & 0 & 0 \\
          0 & 0 & 0 & \sigma & 0 \\
          0 & 0 & 0 & 0 & \sigma\\
          \end{bmatrix}
    \end{equation}\
-   Observational error is the same for all functional groups and is assumed to be 0 (we assume complete knowledge of the system): $v_t = v$, thus `R = 0`
-   Constant covariates\*: $C_{t} = C$ and $c_{t} = c = 0$
-   We assume that the system is at equillibrium, therefore: U = A = 0
-   $X_0 = X_{t=1}$, which is passed to the `MARSS()` as the argument `tinitx = 1`

Some of these assumptions are required to calculate certain multivariate stability metrics.
Thus, calculation of asymptotic resilience that is used here and that follows @downing_temporal_2020, requires that process error is zero and that covariates are constant.

Such model is formalized as:

```{=tex}
\begin{equation}
X = \mathbf{B} \times X_{t-1} + C.0 + 0 + 0 \\

Y =  Z  \times X_t + 0 + D_t.0 + v \\
\end{equation}
```

Thus, our [aquatic commmunity data](aquatic-data) must be split by the concentration of the pesticide, into matrices $Y_{control}, Y_{0.1}, Y_{0.9}, Y_6, Y_{44}$, each containing the time-series of functional group counts under one treatment.
We have two replicates per treatment and four replicates for the control but we are modelling one observation as corresponding to one state processes, therefore a choice must be made how to deal with this. One possibility is to randomly pick one of the replicates per treatment. Alternatively, one can choose to average the data over replicates. 
We here use the mean of species abundances over all replicates:

```{r collapse=TRUE}
## calculate z-scores of abundances
aquacommz_allscen.ldf <- aquacomm_fgps %>%
    dplyr::filter(time >= 1 , time <= 28) %>%
    ungroup() %>%
    dplyr::mutate_at(vars(herb, detr_herb, carn, omni, detr), 
                     ~MARSS::zscore(.)) %>%
    dplyr::mutate(across(c(herb, detr_herb, carn, omni, detr),
                         ~dplyr::na_if(., 0))) %>%
    tidyr::pivot_longer(cols = c(herb, detr_herb, carn, omni, detr),
                        names_to = "fgp",
                        values_to = "abund_z")

## summarize abunds over replicates
aquacommz_allscen.summldf <- aquacommz_allscen.ldf %>%
    dplyr::group_by(time, treat, fgp) %>%
    dplyr::summarize(abundz_mu = mean (abund_z),
                     abundz_sd = sd(abund_z)) %>%
    dplyr::ungroup()

## convert into time-series matrix
aquacommz_allscen.summmxls <- aquacommz_allscen.summldf %>%
    dplyr::select(time, treat, fgp, abundz_mu) %>%
    split(.$treat) %>%
    map(~ dplyr::select(., time, fgp, abundz_mu) %>%
        unique() %>%
        tidyr::pivot_wider(id_cols = fgp,
                           names_from = time, values_from = "abundz_mu", 
                           names_prefix = "time ") %>%
        tibble::column_to_rownames(var = "fgp") %>%
        as.matrix())
```

Next, we prepare the arguments to be used for specifying the MARSS model according to the minimal requirements case outlined above.
As we want to estimate all interaction strengths, we set `B = "unconstrained"`.

```{r cache = TRUE, collapse=TRUE}
## 5 groups
### no process variance, 5 gps
Q_05 <- matrix(list(0), 5, 5)
## single process variance, no covariance
Q_I5 <- matrix(list(0), 5, 5)
diag(Q_I5) <- rep("var_q", 5)
### no observation error variance: 4 reps, 5 gps
R_0r5 <- matrix(list(0), 4*5, 4*5)
### no observation error variance: 0 reps (summarized data), 5 gps
R_05 <- matrix(list(0), 5, 5)
### single error variancee: 0 reps (summarized data), 5 gps
R_I5 <- matrix(list(0), 5, 5)
diag(R_I5) <- rep("var_r", 5)
### single error variancee: 4 reps, 5 gps
R_Ir5 <- matrix(list(0), 4*5, 4*5)
diag(R_Ir5) <- rep("var_r", 4*5)
### single replicate/observation per process: 1 reps, 5 gps
Z_I5 <- matrix(list(0), 5, 5)
diag(Z_I5) <- 1
### multiple replicate/observation per process
Z_r5 <- build_Zreps(4, 5)

aquacommz_allscen.marssls <- aquacommz_allscen.summmxls %>%
    map(~ MARSS(.,
                list(B = "unconstrained",
                     U = "zero", A = "zero",
                     Z = "identity", ## Z_I5 could also have been provided here 
                     Q = "diagonal and equal", 
                     R = R_05,
                     tinitx = 1),
                method = "BFGS"))
names(aquacommz_allscen.marssls) <- paste0("Conc. = ", c("0", "0.1", "0.9", "6", "44" ), " micro g/L")
```

We provide the function `estar::extractB()` which extracts $\mathbf{B}$ in matrix form from the fitted MARSS object and names the interacting groups

```{r}
(aquacomm.Bls <- aquacommz_allscen.marssls %>%
  map(~estar::extractB(., 
                states_names = c("Herbivores", "DetHerbivores", "Carnivores", "Omnivores", "Detrivores"))))
```

Finally, having estimated $\mathbf{B}$ we can calculate all the multivariate stability metrics:

## Reactivity

A positive value indicates the system is reactive and the higher its value, the faster the perturbation can grow. The aquatic community in our example seems to be reactive even when no insecticide is applied.

```{r}
aquacomm.Bls %>%
  map(estar::reactivity)
```

## Maximal amplification

The largest response of the community to the perturbation, it is a measure of the transient behaviour of the system.

```{r}
aquacomm.Bls %>%
  map(estar::max_amp)
```

## Initial resilience

Initial resilience is the initial rate of return to equilibrium. The larger its value, the more stable the system, as its “worst case” initial rate of return to equilibrium is faster (Downing et al. 2020).

```{r}
aquacomm.Bls %>%
  map(estar::init_resil)
```

## Asymptotic resilience

The slowest/long-term asymptotic rate of return to equilibrium after a pulse perturbation. Asymptotic resilience can take on positive real numbers. The larger its value, the more stable the system as such communities have a faster long-term return to stable state.

```{r}
aquacomm.Bls %>%
  map(estar::asympt_resil)
```

## Stochastic invariability

Inverse of the maximal response variance to white noise. The larger its value, the more stable the system as it indicates less variability due to random displacements of the system from the equilibrium.

```{r}
aquacomm.Bls %>%
  map(estar::stoch_var)
```

# References

--------------------------------------------------------------------------------

<details>

<summary>Session Info</summary>

```{r sessionInfo}
Sys.time()
sessionInfo()
```

</details>

<br>

