---
title: "Multivariate Autoregressive State-Space Models"
header-includes:
  \usepackage{amsmath}
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{marss_models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knit-setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE
)
rmarkdown::html_document(df_print = knitr::kable)
```

```{r message=FALSE}
##library(eStar)
library(tidyverse)
library(viridis)
library(kableExtra)
library(imputeTS)
library(MARSS)
library(cowplot)
```

# Introduction to MARSS models

In the `estar` package, we focus on applying a multivariate autoregressive state-space (MARSS) models to estimate the interaction strengths between $m$ species composing a community for which we have abundance data over time $T$. 
To facilitate understanding, we illustrate the definition of a MARSS model in the context of this application.
For a more general and formal definition, refer to the [`MARSS` package documentation](https://nwfsc-timeseries.github.io/MARSS/).

A complete (MARSS) model that describes a state process ($\mathbf{X}$) assessed via an observation process ($\mathbf{Y}$) is defined as:
\begin{equation}
\begin{gathered}
\mathbf{x}_t = \mathbf{B}_{t}\mathbf{x}_{t-1}+\mathbf{u}+\mathbf{C}_{t}\mathbf{c}_{t}+\mathbf{w}_t \text{ where } \mathbf{w}_t \sim \ \text{MVN}(0,\mathbf{Q}) \\
\mathbf{y}_t = \mathbf{Z}_{t}\mathbf{x}_t+\mathbf{a}+\mathbf{D}_{t}\mathbf{d}_{t}+\mathbf{v}_t \text{ where } \mathbf{v}_t \sim \ \text{MVN}(0,\mathbf{R}) \\
\mathbf{x}_1 \sim\ \text{MVN}(\pi,\Lambda) \text{ or } \mathbf{x}_0 \sim\ \text{MVN}(\pi,\Lambda)
\end{gathered}
\end{equation}

In our study case:

- $\mathbf{X}$ is a $m \times T$ matrix, composed of the unknown real abundance values of $m$ species (rows), over time $T$ (columns)
- $\mathbf{Y}$ is a $n \times T$ matrix, composed of $n$ observations of $m$ species (rows, $n \geq m$) over time $T$ (columns) 

The state process ($\mathbf{X}$) is composed by:

- $\mathbf{x}_t$ is a $m \times 1$ vector of unkown populations abundances of the $m$ species at time $t$.
- $\mathbf{B}_t$ is a $m \times m$ matrix of states covariance. This is the interaction matrix of interactions strengths we want to estimate for our data.
- $\mathbf{u}$ is a $m \times 1$ vector of states 
- $\mathbf{C}_t$is a $m \times p$ matrix
- $\mathbf{c}_t$ is a $p \times 1$ matrix
- $\mathbf{w}_t$ is a column vector of process errors, which follow a multivariate normal distribution of mean 0, and covariance $\mathbf{Q}_t$ (a $m \times m$ matrix)

In the observation process ($\mathbf{Y}$):

- $\mathbf{y}_t$ is a column vector of length $n$, with $n$ being the number of observations taken at time $t$.
- $\mathbf{Z}_t$ is a $\mathbf{n} \times m$ matrix which describes the relationship between $\mathbf{x}$ ($m \times 1$ vector) and $\mathbf{y}$ ($n \times 1$ vector).
- $\mathbf{a}$ is a column vector of length $n$
- $\mathbf{D}_t$ is a $n \times q$ matrix
- $\mathbf{d}_t$ s a $q \times 1$ matrix
- $\mathbf{v}_t$ is a column vector of observation errors, which follow a multivariate normal distribution of mean 0, and covariance $\mathbf{R}_t$ (a $n \times n$ matrix) .

# Process and observations errors

autoregressive and non-independent error can be included in B1


# Data overview

## perfect scenario:

```{r}
load("../../data_prep/windblow_barkbeetle/windthrow_longts.RData")
```

```{r eval=FALSE}
pllongts_raw.forestlist[[1]] %>% 
  t %>%
  cbind("t" = as.integer(rownames(.)),.) %>%
  as_tibble() 
```

```{r}
plot_raw <- function(df){
  df %>%
  ggplot(aes(x = time, y = cover, colour = gp, group = gp)) +
  geom_line() +
  geom_point() +
  facet_wrap(~rep) +
  scale_colour_manual(values = viridis(n = 5)) +
  labs(x = "Time", y = "Cover (%)", colour = "Func.\ngroup") +
  theme_minimal()
}

pllongts_raw.forestdf %>%
  ## convert time to absolute, instead of time since disturbance
  dplyr::mutate(time = d_time + abs(min(d_time))) %>%
  tidyr::pivot_longer(col = gp1:gp5,
                      names_to = "gp",
                      values_to = "cover") %>%
  plot_raw()
```

```{r}
pllongts_raw.barkdf %>%
  dplyr::rename(time = d_time) %>%
  tidyr::pivot_longer(col = gp1:gp5,
                      names_to = "gp",
                      values_to = "cover") %>%
  plot_raw() +
  geom_vline(xintercept = 0, color = "#ff0066")
```

MARSS takes y  s a  matrix ts object```{r}

```

