---
title: "eStar package" 
author: "Ludmilla Figueiredo"           
date: "`r Sys.Date()`"                  
output:
  rmdformats::readthedown:
    code_folding: hide                  
    toc_depth: 3                        
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = TRUE,
                      fig.width = 9, fig.height = 6, dpi = 500, 
                      retina = 1, fig.showtext = TRUE)
library(tidyverse)
```


This is an open lab notebook, containing all relevant pieces of information regarding the development of this package.
Information here will likely overlap with the package documentation (and accompanying manuscript), and the later have precedence, since they are the official documentation.

***
# Toy data set
```{r toy-data, fig.cap="Values of the state variable (stat_var) in the baseline and disturbed time-series of the toy data set."}
load("../data/toy_blts.rda")
load("../data/toy_svts.rda")
bind_rows(dplyr::mutate(toy_blts, type = "baseline"),
           dplyr::mutate(toy_svts, type = "disturbed")) %>%
  ggplot(aes(x = time, y = stat_var, colour = type, group = type)) +
  geom_line()+
  geom_point()+
  scale_color_viridis_d(option = "cividis")+
  theme_minimal()
```


# Glossary

# Some tests of functions
```{r eval=FALSE}
check_inv <- invariability(svts_path = './data-raw/toyset.csv', variant = 'cv') 
## important: takes the whole time series (but it is possible that some times steps prior to the distrubace are also included in the provided time series)
mean(data$sv) / sd(data$sv)

## simple test for NA presence
data$sv[18] <- NA
write.csv(data, './data-raw/toyset.csv')

## check NAs
check_inv_NA <- invariability(svts_path = './data-raw/toyset.csv', variant = 'cv') 
check_inv_NA  ## here you see that will cause a pb


## check with the baseline
data_base <- data.frame(sv = rnorm(20, 5, 1), t = 1:20)
write.csv(data_base, './data-raw/toyset_base.csv')
check_inv_base <- invariability(svts_path = './data-raw/toyset.csv', variant = 'bla',
                                time_frame = c(1, 20), bl_path = './data-raw/toysetc_base.csv') ## not sure what should be indicated as a variant here 

```

# Input formats 

<details><summary>Session Info</summary>

```{r sessionInfo}
## DO NOT REMOVE!
## We store the settings of your computer and the current versions of the
## packages used to allow for reproducibility
Sys.time()
#git2r::repository() ## uncomment if you are using GitHub
sessionInfo()
```

</details>

